---
title: "Supplementary note: simulation with nanoSim"
output: github_document
---

```{r setup, include=FALSE}
library(here)
library(tidyverse)
knitr::opts_chunk$set(root.dir = here(), comment = NA, echo = FALSE, message = FALSE, warnings = FALSE, include = TRUE, fig.align = 'center',fig.retina=2)
```

We implemented a small in-silico simulation with NanoSim from synthetic tumour genome, to provide a controlled stress-test of somatic variant calling (SNVs) across tumour purity and sequencing depth while keeping the main benchmarking focused on empirical ONT mixture datasets.

# Overview

## Goal
This simulation is designed as a *supplementary sanity check* (orthogonal to the real mixture benchmarking) to isolate expected failure modes as a function of:

- Tumour purity (fraction of tumour reads in the mixture)
- Sequencing depth (15×–60× on chr22)

We simulate **matched tumour/normal** ONT datasets on a **single chromosome (chr22)** with:
- A shared germline background present in both tumour and normal
- Additional somatic SNVs present only in tumour (the ground truth)


## Inputs
- `chr22.fasta` from GRCh38 (reference genome FASTA; indexed)
- A real ONT FASTQ for NanoSim training, COLO829_BL.bam
- Synthetic germline genome and tumour genome for COLO829_BL and COLO829.

**Tools:**

 - `SAMtools`
 - `BCFtools`
 - `NanoSim`
 - `cstag`
 - `chopper`
 - `minimap2`
 - `clairS`
 - `DeepSomatic`
 
See detailed steps in the [Snakefile](../Snakefile)

## Bechmarking results

```{r}
df <- list.files("../benchmark", pattern = "summary.txt", full.names = TRUE, recursive = TRUE) |> 
  set_names(\(f) dirname(f) |> str_remove("../benchmark/")) |> 
  map(\(x) read.table(x,header = TRUE)) |> 
  list_rbind(names_to = "prefix") |> 
  separate_wider_delim(prefix, "/", names = c("tool", "combination")) |> 
  separate_wider_delim(combination, ".", names=c("sample", "depth_t", "normal", "depth_n")) |> 
  mutate(purity=str_split_i(sample, "_",i=2) |> as.integer()) |> 
  select(tool, sample, purity, depth_t, depth_n, Precision, Recall, F1.score) |> 
  arrange(tool, desc(purity), desc(depth_t), desc(depth_n)) |> 
  rename("Tool"=tool, "Sample"=sample, "Tumour purity"=purity, "Tumour depth"=depth_t, "Normal depth"=depth_n) 

df |> knitr::kable()
df |> write_excel_csv("Supplementary Table 3.csv")
```
