---
title: "Germline leakage"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(here)
library(ggh4x)
public_tool_names = c("clairS" = "ClairS", "deepsomatic" = "DeepSomatic", "delly"="Delly","nanomonsv" = "nanomonsv", "savana" = "SAVANA", "severus"= "Severus")
knitr::opts_chunk$set(root.dir = here(), comment = NA, echo = FALSE, message = FALSE, warnings = FALSE, include = TRUE, fig.align = 'center', fig.retina=2)
```

Somatic variant detection pipelines can mistakenly identify germline variants as somatic ones, a process called “germline leakage”. Here I evaluate how germline leakage affected by tumour purity and sequencing depth. To estimate the germline leakage rate, calculate the proportion of SNVs in the somatic calls with a max population frequency > 0.1 in the gnomAD database. Note that this is just a rough estimate which help me to compare across conditions.

```{r, fig.width=6, fig.height=3}
cols<- c("chrom","pos","ref","alt","qual","vaf","depth","freq")

leakage_purity <- list.files(path = "~/working/bioprojects/nanopore_celllines_benchmark/1.dna_mixing_celllines/work/analysis/snvs", full.names = TRUE, recursive = TRUE, pattern = "^snv_gnomad_af_anno.tsv$" ) |> 
  map(\(x) read_tsv(x, col_names=cols) |> add_column(prefix=x)) |> list_rbind() |> mutate(sample=str_split_i(prefix, "/",i=16), tool=str_split_i(prefix,"/",i=13)) |> 
  separate_wider_delim(sample,delim = ".", names = c("tumour","normal")) |> select(-prefix) |> 
  filter(!(chrom=="chrX" & normal=="HCC1937_BL")) |> 
  group_by(tumour,tool) |> summarise(n=n(), prop=mean(freq > 0.1, na.rm = TRUE)) |> ungroup() |> 
  mutate(celltype=ifelse(startsWith(tumour,"COLO829"),"COLO829","HCC1937"), purity=str_split_i(tumour,"_",i=2), purity=ifelse(is.na(purity),"100",purity), purity=factor(purity,levels=seq(100,10,-10)))


leakage_purity  |> ggplot(aes(x=purity,y=prop,group=celltype))  + 
  geom_line(color="darkgrey", linewidth=0.8) +
  geom_point(color="darkblue") +
  facet_grid(celltype~tool, scale="free_y", labeller = labeller(tool=public_tool_names))  + 
  labs(x="Tumour Purity",y="Germline leakage rate", color="")+ 
  theme_bw() + 
  theme(strip.background = element_rect(fill="white"), legend.position = "bottom")

ggsave("figures/supp-figureS17a.pdf", width = 6, height = 3, dpi = "retina")
```
Seems the tumour purity didn't impact much in terms of germline leakage. However, for sample with less then 30% tumour purity, ClairS calls had increased germline leakage whereas DeepSomatic showed decreased germline leakage rate.

**How about sequencing depth?**

```{r, fig.width=6, fig.height=3.2}
leakage_depth <- list.files(path = "~/working/bioprojects/nanopore_celllines_benchmark/2.simulate_sequencing_depth/analysis/snvs", full.names = TRUE, recursive = TRUE, pattern = "^snv_gnomad_af_anno.tsv$" ) |> 
  map(\(x) read_tsv(x, col_names=cols) |> add_column(prefix=x)) |> list_rbind() |> mutate(sample=str_split_i(prefix, "/",i=13), tool=str_split_i(prefix,"/",i=12)) |> 
  separate_wider_delim(sample,delim = ".", names = c("tumour","t_depth","normal","n_depth")) |> select(-prefix) |> 
  filter(!(chrom=="chrX" & normal=="HCC1937_BL")) |> 
  group_by(tumour,t_depth, n_depth, tool,) |> summarise(n=n(), prop=mean(freq > 0.1, na.rm = TRUE)) |> ungroup() |> 
  mutate(celltype=ifelse(startsWith(tumour,"COLO829"),"COLO829","HCC1937"), purity=str_split_i(tumour,"_",i=2), purity=ifelse(is.na(purity),"100",purity), purity=factor(purity,levels=seq(100,10,-10)))

leakage_depth |> filter(purity=="100") |> 
  unite(comb,c(t_depth,n_depth),sep="-") |> mutate(comb=factor(comb, levels=unique(comb) |> rev())) |> 
  ggplot(aes(x=comb,y=prop,group=celltype, color=celltype)) +
  geom_line(color="darkgrey", linewidth=0.8) +
  geom_point(color="darkblue") +
  facet_grid2(celltype~tool, scales = "free_y", labeller = labeller(tool=public_tool_names), independent = "y") +
  labs(x="Read depth (tumour-normal)",y="Germline leakage rate", color="")+ 
  theme_bw() + 
  theme(strip.background = element_rect(fill="white"), legend.position = "bottom", axis.text.x = element_text(angle=45, hjust = 1))

ggsave("figures/supp-figureS17b.pdf", width = 6, height = 3.2, dpi = "retina")
```

The germline leakage rate spiked when germline sample sequenced at 15x depth.
